---
title: "Major Weather Events from US NOAA Data"
author: "Haonan Xu"
date: "8/22/2020"
output: html_document
---
### Synopsis

******

### Background
The [NOAA Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) can be downloaded here.

### Data Processing
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Reading Data

```{r LoadPackage, message=FALSE}
library(tidyverse)
library(magrittr)
library(lattice)
library(knitr)
library(lubridate)
```

This portion of the code downloads the NOAA data and processes it

```{r ReadData, cache=TRUE}
filelink <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

download.file(filelink,"StormData.csv.bz2", method = "curl")

stormdata <- read.csv("StormData.csv.bz2")

stormdata2 <- stormdata
stormdata2 <- stormdata2 %>%
  mutate(Time = sub(" 0:00:00", "", BGN_DATE)) %>%
  mutate(Time = as.POSIXct(Time,format = "%m/%d/%Y")) %>%
  select(Time,TIME_ZONE,STATE,EVTYPE,FATALITIES,INJURIES,
                                   PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP) %>%
  filter(!grepl("summary",stormdata2$EVTYPE,ignore.case = T)) %>%
  mutate(EVTYPE = as.factor(EVTYPE))

remove(stormdata) # Original data is removed for memory efficiency
```

#### Data Processing
The following section processes the storm data for calculations
```{r}
## Data contain summaries as observations, i.e. obs. 278098 is "Summary: Nov. 16"
## The filter() function above remove these observations 

exponents <- unique(c(stormdata2$PROPDMGEXP,stormdata2$CROPDMGEXP))
exp_value <- 10**c(3,6,0,9,6,NA,0,5,6,NA,4,2,3,NA,7,NA,NA,1,8,3)

## Exponents of the economic damage are noted as "K", "M", "B", etc. 
## The above lines converts it to numeric representations 

stormdata2$CROPDMGEXP[stormdata2$CROPDMGEXP %in% exponents] <- exp_value[match(
  stormdata2$CROPDMGEXP,exponents, nomatch = NA
)]

stormdata2$PROPDMGEXP[stormdata2$PROPDMGEXP %in% exponents] <- exp_value[match(
  stormdata2$PROPDMGEXP,exponents, nomatch = NA
)]

stormdata2$PROPDMGEXP <- as.numeric(stormdata2$PROPDMGEXP)
stormdata2$CROPDMGEXP <- as.numeric(stormdata2$CROPDMGEXP)

stormdata2 <- stormdata2 %>% 
  mutate(PROPTOTAL = PROPDMG * PROPDMGEXP/10**6,
         CROPTOTAL = CROPDMG * CROPDMGEXP/10**6, DAMAGETOTAL = CROPTOTAL + PROPTOTAL,
         CASUALTY = FATALITIES + INJURIES, YEAR = year(Time))

## unknown exponents like "+", "?", "h", "H", or "-" were also introduced in the data
## For the purpose of this study, they are assumed to be NA 
## k, K == 1000; m, M == 10**6; B == 10**9; 7 == 10** 7 etc

```

The following section calculates the casualty by weather events
```{r Casualty}
casualty <- stormdata2 %>% group_by(EVTYPE) %>%
  summarise(count = n(), Total_Fatality = sum(FATALITIES), 
            Total_Injury = sum(INJURIES), Mean_Fatality = mean(FATALITIES),
            Mean_injury = mean(INJURIES),
            Total = Total_Fatality + Total_Injury,
            .groups = "drop_last") %>%
  ungroup() %>%
  arrange(desc(Total))

```

The following section calculates the damage by weather events
```{r Damage}

## Damage is recorded in millions of dollars 
damage <- stormdata2 %>% group_by(EVTYPE) %>%
  summarise(count = n(), Total_Property = sum(PROPTOTAL),
            Total_Crop = sum(CROPTOTAL), Mean_Property = mean(PROPTOTAL),
            Mean_Crop = mean(CROPTOTAL), Total = Total_Property + Total_Crop,
            .groups = "drop_last") %>%
  ungroup() %>%
  arrange(desc(Total))
```
******


#### Results
In the U.S., the weather event most harmful to public health is **`r casualty$EVTYPE[1]`**, and the one with greatest economic consequences is **`r damage$EVTYPE[1]`**. To compute for the threat to public health, the total fatalities and total injuries are considered; for assessing economic damage, total property damage and total crop damage are added. 

```{r table}
casualty_table <- head(casualty,3) %>%
  mutate(Ranking = row_number(), Event = EVTYPE) %>%
  select(Ranking, Event, Total_Injury, Total_Fatality, Total)


damage_table <- head(damage,3) %>%
  mutate(Ranking = row_number(), Event = EVTYPE) %>%
  select(Ranking, Event, Total_Property, Total_Crop,Total)

#Formatting the table by reducing line length
damage_table$Event <- gsub("/TYPHOON", "/ TYPHOON", damage_table$Event)

kables(list(
  kable(casualty_table, align = "c", caption = "Table 1. Weather Events with the most Casualty"),
  kable(damage_table, align = "c", caption = "Table 2. Weather Events with the most Economic Damage ($ Mil)")
))

```

```{r Lattice}

plotting <- stormdata2 %>% group_by(EVTYPE) %>%
  mutate(Event_Count = n()) %>%
  arrange(desc(Event_Count))

most_freq <- as.character(unique(plotting$EVTYPE)[1:20])

plotting <- plotting %>%
  filter(EVTYPE %in% most_freq) %>%
  mutate(EVTYPE = droplevels(EVTYPE)) %>%
  select(Time, YEAR, STATE, EVTYPE, CASUALTY, DAMAGETOTAL)

plotting <- plotting %>%
  group_by(EVTYPE,YEAR) %>%
  summarise(EVCOUNT = n(), CASUALTY_Y = sum(CASUALTY,na.rm = T), 
            DAMAGE_Y = sum(DAMAGETOTAL,na.rm = T),
            .groups = "drop_last") %>%
  ungroup()

xyplot(CASUALTY_Y ~ YEAR | EVTYPE, plotting, type = "l",
       par.strip.text = list(cex = 0.4),
       ylab = "Total Annual Casualty (person)")
xyplot(DAMAGE_Y ~ YEAR | EVTYPE, plotting,type = "l",
       par.strip.text = list(cex = 0.4),
       ylab = "Total Annual Damage (Millions USD) ")
  
```

```{r Histogram}

par(mfrow = c(2,2))
flood <- stormdata2[which(stormdata2$EVTYPE == "FLOOD"),]
hist(flood$Time, freq = T, breaks = 12, main = "Histogram of Flood Frequency",
     xlab = "Time")

tornado <- stormdata2[which(stormdata2$EVTYPE == "TORNADO"),]
tornado <- mutate(tornado, casualty = FATALITIES + INJURIES)
hist(tornado$Time, freq = T, breaks = 12, main = "Histogram of Tornado Frequency",
     xlab = "Time")

```

